Quantum Mechanics Assignment Seven

### 2016-02-24 quantum/latex/a7.tex

\title{Quantum Mechanics Assignment Six}
\author{Elliott Capek}
\titlespacing{\section}{0pt}{0pt}{5pt} % Spacing around titles

@@

\section{11.12: blah}

@@

The labels are kind of like spectroscopic notation. The upper term (3d, 2p, etc...) is
the n and $\ell$ number for the state, where $\ell$ is translated into alphabet
notation. The subscript is the

>>>>

\title{Quantum Mechanics Assignment Seven}
\author{Elliott Capek}
\titlespacing{\section}{0pt}{0pt}{5pt} % Spacing around titles

@@

\section{Problem 1: Relativistic Fine Correction}

This problem asks us to work through parts of the derivation of perturbation theory applied to
the relativistic correction to the Hydrogen energy spectrum.\\

\textbf{a.)}Here we show how to find the expectation value of $p^4$ more easily:\\

\begin{align*}
  &\bra{n,\ell,m}p^4\ket{n,\ell,m}\\
  &= \bra{n,\ell,m}p^2p^2\ket{n,\ell,m}\\
  &= 4m^2\bra{n,\ell,m}\left(H_0 + \frac{e^2}{4\pi\epsilon r}\right)
  \left(H_0 + \frac{e^2}{4\pi\epsilon r}\right)\ket{n,\ell,m}\\
  &= 4m^2\left(\bra{n,\ell,m}H_0^2\ket{n,\ell,m}
  + \bra{n,\ell,m}H_0\frac{e^2}{4\pi\epsilon r}\ket{n,\ell,m}
  + \bra{n,\ell,m}\frac{e^2}{4\pi\epsilon r}H_0\ket{n,\ell,m}
  + \bra{n,\ell,m}\frac{e^4}{16\pi^2\epsilon^2r^2}\ket{n,\ell,m}\right)\\
  &= 4m^2\left((E_n^{(0)})^2
  + E_n^{(0)}\bra{n,\ell,m}\frac{e^2}{4\pi\epsilon r}\ket{n,\ell,m}
  + \bra{n,\ell,m}\frac{e^2}{4\pi\epsilon r}\ket{n,\ell,m}E_n^{(0)}
  + \frac{e^4}{16\pi^2\epsilon^2r^2}\bra{n,\ell,m}\frac{1}{r^2}\ket{n,\ell,m}\right)\\
  &= 4m^2\left((E_n^{(0)})^2
  + \frac{e^2E_n^{(0)}}{4\pi\epsilon}\braket{\frac{1}{r}}
  + \frac{e^2}{4\pi\epsilon}\braket{\frac{1}{r}}E_n^{(0)}
  + \frac{e^4}{16\pi^2\epsilon^2r^2}\braket{\frac{1}{r^2}}\right)\\
  &= 4m^2\left((E_n^{(0)})^2
  + \frac{e^2E_n^{(0)}}{2\pi\epsilon}\braket{\frac{1}{r}}
  + \frac{e^4}{16\pi^2\epsilon^2r^2}\braket{\frac{1}{r^2}}\right)\\
\end{align*}

\textbf{b.)}\\

\textbf{c.)} We perform the following integrals using Mathematica:

\begin{align*}
  \braket{\frac{1}{r}}_{20} &= \left(\frac{Z}{2a}\right)^3\int_0^\infty
  4r\left(1-\frac{Zr}{2a}\right)^2e^{-Zr/a}dr = \frac{Z}{4a}\\
  \braket{\frac{1}{r^2}}_{20} &= \left(\frac{Z}{2a}\right)^3\int_0^\infty
  4\left(1-\frac{Zr}{2a}\right)^2e^{-Zr/a}dr = \frac{Z^2}{4a^2}\\
  \braket{\frac{1}{r}}_{21} &= \left(\frac{Z}{2a}\right)^3\int_0^\infty
  \frac{Z^2r}{3a^2}e^{Zr/a}dr = \frac{Z}{4a}\\
  \braket{\frac{1}{r^2}}_{21} &= \left(\frac{Z^5}{24a^5}\right)^3\int_0^\infty
  r^2e^{-Zr/a}dr = \frac{Z^2}{12a^2}\\
\end{align*}

These are the desired results. We then assume this is true for all n, $\ell$ quantum numbers.\\

@@

The labels are kind of like spectroscopic notation. The upper term (3d, 2p, etc...) is
the n and $\ell$ number for the state, where $\ell$ is translated into alphabet
notation. The subscript is the F-number (L+S) for that orbital. The spin of the electron can be
$\pm\frac{1}{2}$, so the F-number plus or minus $\frac{1}{2}$ is the $\ell$-number. $3d_{5/2}$ is
the n=3, $\ell$=2 orbital \textit{where the angular momentum and spin momentum combine}. There is
another orbital where they are subtracted.\\

The $3p_{3/2}$ and $3d_{3/2}$ orbitals both have corrections of 1.62GHz, as seen in the table. They
are made up of the following linear combinations of uncoupled states:\\

\begin{align*}
  \ket{n,F,M_F} &= \ket{n,M_I,M_s}\\
  \ket{3,\frac{3}{2}, \frac{ 3}{2}} &= \ket{3,1,\frac{1}{2}}\\
  \ket{3,\frac{3}{2}, \frac{ 1}{2}} &= \frac{1}{\sqrt{3}}\ket{3,1,\frac{1}{2}}
  + \sqrt{\frac{2}{3}}\ket{2,0,\frac{1}{2}}\\
  \ket{3,\frac{3}{2}, \frac{-1}{2}} &= \sqrt{\frac{2}{3}}\ket{3,0,-\frac{1}{2}}
  + \frac{1}{\sqrt{3}}\ket{3,-1,\frac{1}{2}}\\
  \ket{3,\frac{3}{2}, \frac{-3}{2}} &= \ket{3,-1,-\frac{1}{2}}\\
\end{align*}

I liked this problem. I didn't get to finish it, but it was interesting thinking about how to use
Clebsch-Gordan tables to compare bases. It was also helpful thinking about how there are different
quantum states for when angular momenta combine vs subtract from each other. This doesn't make
sense but is an important part of quantum mechanics.\\
